---
title: "treestructure applied to structured coalescent simulation"
author: "Erik Volz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{structuredCoalescent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=11
)
```


# `treestructure`

## Citation


Identification of hidden population structure in time-scaled phylogenies

Erik Volz, Carsten Wiuf, Yonatan Grad, Simon Frost, Ann Dennis, Xavier Didelot,
bioRxiv 704528; [https://doi.org/10.1101/704528](https://doi.org/10.1101/704528)

## Structured coalescent simulation 

This example shows `trestruct` applied to a simulated structured coalescent tree that includes samples from a large constant size population and samples from three small 'outbreaks' which are growing exponentially. These simulations were generated with the `phydynR` [package](https://github.com/emvolz-phylodynamics/phydynR). 

```{r message=FALSE}
library(treestructure)
```

Load the tree: 

```{r}
( tree <- ape::read.tree( system.file('sim.nwk', package='treestructure') ) ) 
```


Note that the tip labels corresponds to the deme of each sample. '1' is the constant size reservoir, and '0' is the exponentially growing deme. 


This will run the treestructure algorithm under default setting: 
```{r message=FALSE}
s <- trestruct( tree ) 
```

You can print the results: 

```{r}
print(s) 
```

The default plotting behavior uses the `ggtree` package if available. 

```{r message=FALSE}
plot(s)  + ggtree::geom_tiplab() 
```

If not, or if desired, `ape` plots are available

```{r}
plot( s, use_ggtree = FALSE )
```


For subsequent analysis, you may want to turn the treestructure result into a dataframe: 
```{r}
structureData <- as.data.frame( s ) 
head( structureData )
```

Each cluster and partition assignment is stored as a factor. You could use `split` to get a data frame for each partition. 
Suppose we want a tree corresponding to partition 1: 
```{r}
with ( structureData, 
	ape::keep.tip(s$tree,  taxon[ partition==1 ] ) 
) -> partition1
partition1
plot(partition1)
```

Two parameters will have large influence on results: 

1. `level` is the significance level for subdividing a clade into a new cluster. To detect more clusters, increase `p`, but note that this will also increase the false positive rate. 
2. `minCladeSize` controls the smallest allowed cluster size in terms of the number of tips. With a smaller value, smaller clusters may be detected, but computation time will increase. 

Example: 
```{r message=FALSE}
trestruct( tree, level = .05, minCladeSize = 5 )
```
