---
title: "Considering node support values on designations and update previous 
    treestructure object with new sequences"
author: "Vinicius Franceschi"
date: "`r Sys.Date()`"
output: 
  bookdown::html_vignette2:
  #rmarkdown::html_vignette
  #bookdown::pdf_book:
    toc: TRUE
pkgdown:
  as_is: true
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Considering node support values on designations and 
      update previous treestructure object with new sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=11
)
```

# Introduction

In this tutorial, we will consider node support values (e.g. bootstrap) on 
cluster designations and update previous `treestructure` object with new sequences.

This tutorial uses SARS-CoV-2 public data available [here](https://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2//2022/04/02/) 
to demonstrate the use of node support values from e.g. parametric bootstrap to 
avoid designating population structure in badly supported clades. 
This tutorial also demonstrates how to update previous cluster designations 
(an existing `treestructure` object) using a new rooted maximum likelihood tree 
incorporating more sequences. The new tips are added to the cluster which shares 
its MRCA (most recent common ancestor).

# Workflow to analyse SARS-CoV-2 sequence data

For a step-by-step guide to replicate the complete workflow please see [here](https://github.com/vinibfranc/demo_treestructure/blob/main/R/prepare.R).

Briefly, we downloaded SARS-CoV-2 public metadata, a treefile, and multiple 
sequence alignments from the UShER index. Then we extracted sequences up to the 
end of February 2020 (n < 1,000 sequences). We then estimated a maximum 
likelihood (ML) tree with [IQ-TREE](http://www.iqtree.org/) v2.2.2.6 with 1,000 
ultrafast bootstraps, a time-scaled tree using 
[`treedater`](https://github.com/emvolz/treedater) (strict clock and 0.0008 subst./site/year), 
removed outliers by root-to-tip regression, 
an then re-estimated the timetree without the outliers (n = 916).


Let's load the resulting ML tree with bootstrap values:
```{r message=FALSE}
library(treestructure)
```


```{r}
mltr2_outl_rm <- readRDS( system.file('mltr2_outl_rm_sc2_feb2020.rds', 
                                      package='treestructure') )

ggtree::ggtree(mltr2_outl_rm)
```

And here you can load the timetree:
```{r}
timetr2_phylo <- readRDS( system.file('timetr2_phylo_sc2_feb2020.rds', 
                                      package='treestructure') )

ggtree::ggtree(timetr2_phylo)
```

Firstly, we will assign clusters without using bootstrap support:
```{r}
trestruct_res_nobt <- trestruct(timetr2_phylo, 
                                minCladeSize = 30, 
                                nodeSupportValues = FALSE, 
                                level = 0.01)

plot(trestruct_res_nobt, use_ggtree = T) + ggtree::geom_tippoint()
```

The `treestructure` analyses resulted in 9 clusters.

Now let's add the support values to a vector that we will pass to `trestruct`:

```{r}
timetr2_boot <- as.integer(mltr2_outl_rm$node.label)
timetr2_boot[is.na(timetr2_boot)] <- 95
print(timetr2_boot)
```

Note that in the above example, some branches did not show a branch support as 
the tree was estimated with IQ-TREE and the alignment contained many identical tips.
In our example we will assign to all these branches without a branch support, a 
node support of 95. However, in your own example, if too many identical sequences 
are present and if you want to estimate the branch support for all branches
when estimating a tree with IQ-TREE, you should use the flag `-keep-ident`. See
[here](https://github.com/iqtree/iqtree2/issues/343) for further details.



And, finally, we designate clusters that have at least 80% bootstrap support:

```{r}
trestruct_res <- trestruct(timetr2_phylo, 
                           minCladeSize = 30, 
                           nodeSupportValues = timetr2_boot, 
                           nodeSupportThreshold = 80, 
                           level = 0.01)
plot(trestruct_res, use_ggtree = T) + ggtree::geom_tippoint()
```

Now we have only 5 well-supported clusters with differences in coalescent patterns.

To update the previous `treestructure` object with new sequences, we extracted 
SARS-CoV-2 sequences up to 15 March 2020 (n > 6,500 sequences). We then estimated 
a new ML tree including all those sequences as before.

Note that this new tree must be rooted, but does not need to be time-scaled or 
binary.

```{r}
mltr_addtips <- readRDS( system.file('mltr_addtips_mar2020.rds', 
                                     package='treestructure') )
ggtree::ggtree(mltr_addtips)
```

And without the need to re-estimate a timetree or re-run `trestruct` from scratch, 
we are now able to add the new sequences to the existing `treestructure` object:

```{r}
trestruct_add_tips <- addtips(trst = trestruct_res, tre = mltr_addtips)
plot(trestruct_add_tips, use_ggtree = T) + ggtree::geom_tippoint()
```
