---
title: "Node support values using treestructure"
author: "Vinicius Franceschi and Fabricia F. Nascimento"
date: "`r Sys.Date()`"
output: 
  bookdown::html_vignette2:
  #rmarkdown::html_vignette
  #bookdown::pdf_book:
    toc: TRUE
bibliography: bib/tutorials.bib
pkgdown:
  as_is: true
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Node support values using treestructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 11,
  warning = FALSE, 
  message = FALSE
)
```

# Introductions

This tutorial uses the public data available for Ebola [here](https://github.com/ebov/space-time) to demonstrate the use of node support 
values (e.g. bootstrap and posterior probability) to avoid designating population 
structure in badly supported clades. 

We will use their [time-scaled phylogenetic tree](https://github.com/ebov/space-time/blob/master/Data/Makona_1610_cds_ig.GLM.MCC.tree) estimated with [BEAST](https://beast.community).


First, we need to load the R package we will use in this tutorial.


```{r message=FALSE}
library(treeio)
library(ggtree)
library(treestructure)
```


Now, we will load the time-scaled phylogenetic tree with posterior probability 
support values:

```{r}
#get the dated tree by first downloading it from the URL below
tree_url <- "https://raw.githubusercontent.com/ebov/space-time/master/Data/Makona_1610_cds_ig.GLM.MCC.tree"
tmp_file <- tempfile(fileext = ".tree")

#Download BEAST tree to a tmp file
download.file(tree_url, tmp_file, mode = "wb")

#read the downloaded tree
beast_tree <- read.beast(tmp_file)
```


Then we need to convert the tree to a `phylo` object and add the posterior 
probability information from the Bayesian tree to the `phylo` object tree.

If we use the `ape:read.nexus` to read the tree that was estimated with BEAST, 
we won't get the posterior probability associated to the clades.

We now convert the dated tree to `phylo` object.

```{r}
dated_tre <- as.phylo(beast_tree)
```

Now we add the posterior probability from the BEAST tree to the dated_tree
object.

```{r}
# Get number of tips
n_tips <- length(dated_tre$tip.label)

# Get BEAST tree as tibble (it will include node numbers and posterior probabilities)
tree_data <- as_tibble(beast_tree)

#get posterior probability
posterior <- as.numeric(tree_data$posterior[(n_tips + 1):nrow(tree_data)])

#add the posterior probability information to the `phylo` object
dated_tre$node.label <- posterior
```



## Assign clusters without using node support

Firstly, we will assign clusters without using node support values. Note that
the parameter `nodeSupportValues` is set to FALSE.

```{r eval=FALSE}
trestruct_res_nobt <- trestruct(dated_tre, 
                                minCladeSize = 30, 
                                nodeSupportValues = FALSE, 
                                level = 0.01)
```

In the above example, the `trestruct` function took 7 minutes to run on a macOS M2. 
Here, we can load the results instead.

```{r}
trestruct_res_nobt <- readRDS( system.file('trestruct_res_nobt.rds',
                                           package='treestructure') )

plot(trestruct_res_nobt, use_ggtree = T) + ggtree::geom_tippoint()
```

The `treestructure` analyses resulted in 23 clusters.


## Assign clusters using branch support

Now, we will assign clusters using the information on branch support.
As we are analyzing a dated tree estimated with BEAST, the branch support is 
the posterior probability.


We designate clusters that have at least 0.95 posterior probability. This
is achieved by setting to 95 the parameter _nodeSupportThreshold_ in the **trestruct**
function. 

Note that now the parameter `nodeSupportValues` is set to TRUE, which tells
the algorithm that the node support values is provided with the `phylo` object
as `node.label`.

You can also provide the `nodeSupportValues` as a vector with length equal to 
the number of internal nodes in the tree.

```{r eval=FALSE}
trestruct_res <- trestruct(dated_tre, 
                           minCladeSize = 30, 
                           nodeSupportValues = TRUE, 
                           nodeSupportThreshold = 95, 
                           level = 0.01)
```

In the above example, the `trestruct` function took 50 seconds to run on a macOS M2. 
Here, we can load the results instead.

```{r}
trestruct_res <- readRDS( system.file('trestruct_res.rds',
                                      package='treestructure') )

plot(trestruct_res, use_ggtree = T) + ggtree::geom_tippoint()
```


Now we have only 6 well-supported clusters with differences in coalescent patterns.

Note that this might change if you use a higher or lower value for the
*nodeSupportThreshold* in the **trestruct** function.


## Using the CH-index 

The CH-index will compare different values for the parameter `levels` in the
`treestructure` algorithm. We will need to specify `level` = NULL and 
specify the lower and upper bound to optimize the parameter `level`.

Note that we will also take into consideration the use of branch support.

```{r eval = FALSE}
trestruct_chindex <- trestruct(dated_tre,
                               minCladeSize = 30, 
                               nodeSupportValues = TRUE, 
                               nodeSupportThreshold = 95,
                               level = NULL,
                               levellb = 0.0001, 
                               levelub = 0.001)
```

In the above example, the `trestruct` function took 5 minutes to run on a macOS M2. 
Here, we can load the results instead.

```{r}
trestruct_chindex <- readRDS( system.file('trestruct_chindex.rds',
                                      package='treestructure') )

plot(trestruct_chindex, use_ggtree = T) + ggtree::geom_tippoint()
```

Now we have only 2 well-supported clusters with differences in coalescent 
patterns after optimizing the significance level.

In this example, instead of using level = 0.01 (as in the default ), the optimization
of level by using the CH-index identified level as 0.00073.


## Combining node support with a false discovery rate

The `level` argument used above is a subjective clustering threshold. Supplying a
target false discovery rate with `fdr` instead controls the whole-tree error rate of
the designation (see the
[false discovery rate](false_discovery_rate.html) vignette). This is complementary
to node support: the false discovery rate asks whether a clade's coalescent pattern
differs significantly, while node support asks whether the clade is topologically
resolved. Both can be required together:

```{r eval=FALSE}
trestruct( dated_tre, fdr = 0.1, minCladeSize = 30,
           nodeSupportValues = TRUE, nodeSupportThreshold = 95 )
```

On this tree the two criteria reinforce one another. With `fdr = 0.1` alone the
method designates four clusters; additionally requiring node support reduces this to
two. The result is insensitive to the exact support threshold, because the clusters
that support removes are designated at nodes with very low posterior probability
(well below 0.5), while those it keeps are strongly supported:

```{r}
readRDS( system.file('vignette_fdr_support.rds', package = 'treestructure') )
```

These two clusters coincide with those selected by the CH index above.


# References
