---
title: "Node support values and update of treestructure object with new sequences"
author: "Vinicius Franceschi"
date: "`r Sys.Date()`"
output: 
  bookdown::html_vignette2:
  #rmarkdown::html_vignette
  #bookdown::pdf_book:
    toc: TRUE
pkgdown:
  as_is: true
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Node support values and update of treestructure object with new sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 11
)
```

# Introduction

In this tutorial, we will consider node support values (e.g. bootstrap) on 
cluster designations and update previous `treestructure` object with new sequences.

This tutorial uses SARS-CoV-2 public data available [here](https://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2//2022/04/02/) 
to demonstrate the use of node support values from e.g. parametric bootstrap to 
avoid designating population structure in badly supported clades. 
This tutorial also demonstrates how to update previous cluster designations 
(an existing `treestructure` object) using a new rooted maximum likelihood tree 
incorporating more sequences. The new tips are added to the cluster which shares 
its MRCA (most recent common ancestor).

# Workflow to analyse SARS-CoV-2 sequence data

For a step-by-step guide to replicate the complete workflow please see [here](https://github.com/vinibfranc/demo_treestructure/blob/main/R/prepare.R).

Briefly, we downloaded SARS-CoV-2 public metadata, a treefile, and multiple 
sequence alignments from the UShER index. Then we extracted sequences up to the 
end of February 2020 (n < 1,000 sequences). We then estimated a maximum 
likelihood (ML) tree with [IQ-TREE](http://www.iqtree.org/) v2.2.2.6 with 1,000 
ultrafast bootstraps, a time-scaled tree using 
[`treedater`](https://github.com/emvolz/treedater) (strict clock and 0.0008 subst./site/year), 
removed outliers by root-to-tip regression, 
an then re-estimated the timetree without the outliers (n = 891).


Let's load the resulting ML tree with bootstrap values:
```{r message=FALSE}
library(treestructure)
```


```{r}
mltr2_outl_rm <- readRDS( system.file('mltr2_outl_rm_sc2_feb2020.rds', 
                                      package='treestructure') )

ggtree::ggtree(mltr2_outl_rm)
```

And here you can load the timetree:
```{r}
timetr2_phylo <- readRDS( system.file('timetr2_phylo_sc2_feb2020.rds', 
                                      package='treestructure') )

ggtree::ggtree(timetr2_phylo)
```


## Assign clusters without using node support

Firstly, we will assign clusters without using bootstrap support:
```{r eval=FALSE}
trestruct_res_nobt <- trestruct(timetr2_phylo, 
                                minCladeSize = 30, 
                                nodeSupportValues = FALSE, 
                                level = 0.01)
```

Because `treestructure` will take several minutes to run, we can load the results:

```{r}
trestruct_res_nobt <- readRDS( system.file('trestruct_res_nobt.rds',
                                           package='treestructure') )

plot(trestruct_res_nobt, use_ggtree = T) + ggtree::geom_tippoint()
```

The `treestructure` analyses resulted in 13 clusters.


## Assign clusters using bootstrap support

Let's add the support values to a vector that we will pass to `trestruct`:

```{r}
timetr2_boot <- as.integer(mltr2_outl_rm$node.label)

#note that IQ-TREE does not give a node support for the "root" of the tree,
#You, as a user with knowledge of your data, will decide if this should be 
#a high or low support value.
timetr2_boot[is.na(timetr2_boot)] <- 95

#show the first 6 node support values.
print(head(timetr2_boot))
```


Finally, we designate clusters that have at least 80% bootstrap support. This
is achieved by setting to 80 the parameter _nodeSupportThreshold_ in the **trestruct**
function.

```{r eval=FALSE}
trestruct_res <- trestruct(timetr2_phylo, 
                           minCladeSize = 30, 
                           nodeSupportValues = timetr2_boot, 
                           nodeSupportThreshold = 80, 
                           level = 0.01)
```

Because it will take a minute to run `treestructure`, we can load the result instead.

```{r}
trestruct_res <- readRDS( system.file('trestruct_res.rds',
                                      package='treestructure') )

plot(trestruct_res, use_ggtree = T) + ggtree::geom_tippoint()
```


Now we have only 4 well-supported clusters with differences in coalescent patterns.
Note that this might change if you use a higher or lower value for the
*nodeSupportThreshold* in the **trestruct** function.


## Update a previous treestrucuture object with new sequences 

To update the previous `treestructure` object with new sequences, we extracted 
SARS-CoV-2 sequences up to 15 March 2020 (n > 6,712 sequences). We then estimated 
a new ML tree including all those sequences as before.

Note that this new tree must be rooted, but does not need to be time-scaled or 
binary.

```{r}
#Note that this tree has more sequences than the previous tree used in this
#tutorial.
mltr_addtips <- readRDS( system.file('mltr_addtips_mar2020.rds', 
                                     package='treestructure') )
ggtree::ggtree(mltr_addtips)
```

And without the need to re-estimate a timetree or re-run `trestruct` from scratch, 
we are now able to add the new sequences to the existing `treestructure` object:

```{r}
trestruct_add_tips <- addtips(trst = trestruct_res, tre = mltr_addtips)
plot(trestruct_add_tips, use_ggtree = T) + ggtree::geom_tippoint()
```


If you would like to compare the sequence names that comprise each cluster in
each tree, you can do:

```{r}

#compare sequences in cluster 1 from trestruct_res object and the 
#trestruct_add_tips object

tree1_cluster1 <- trestruct_res$clusterSets$`1`
tree2_cluster1 <- trestruct_add_tips$clusterSets$`1`

length(tree1_cluster1)
length(tree2_cluster1)

```

Note that the length of tree1_cluster1 and tree2_cluster1 is different.
That is because we _added_ tips from the ML tree, _mltr_addtips_, to the 
`treestructure` object, _trestruct_res_.

You can also see that all elements in tree1_cluster1 is contained in tree2_cluster1

```{r}

sum(tree1_cluster1 %in% tree2_cluster1)

```
